Seminar: Programming models and Code Generation\\ Winter Term 2014/2015\\ {\bf Quasi deterministic parallel programming using LVars}

### 2015-01-03 draft.tex
New

The purpose of a deterministic-by-construction programming model is to rule out programs like the following from being written, or force them to behave deterministically. \cite{lindsey}Written in a hypothetical parallel language. 
\begin{verbatim}
let _ = put lv 3 in  /write 3 in lv/
let par v = get lv  /read from lv/
        _ = put lv 4  /write 4 in lv/
	      in v   /final value is v/
\end{verbatim} 
The \textit{let par} expression which has \textit{fork-join semantics}, evaluates the two sub-expressions in parallel. The value of $v$ is the value of the entire program. Depending on the scheduler \textit{get} $lv$ or \textit{put} $lv\ 4$ can run first, $v$ might be either $3$ or $4$ during multiple runs giving rise to non determinism.\par
A popular approach for enforcing determinism is to require that variables can only be assigned to once, giving us a single-assignment language. Our program would be invalid because it tries to write to $lv$ twice. Single-assignment variables with blocking read semantics, often known as IVars, turn up in all kinds of deterministic parallel systems. But the single-assignment rule disallows a number of programs that we would like to permit such as writing the same value to an IVar twice. As another example, consider the problem of filling in various parts of a data structure in parallel. \cite{lindsey}Say that $lv$ contains a pair, and we are writing to $lv$ twice, with each write filling in one component of the pair. Please note that $x$ is equal to $\bot$ below.\par 
\begin{verbatim}
let par  _ = put lv (4,x)
         _ = put lv (x,3) /error/
         in get lv
\end{verbatim}  
Read from $lv$ is possible once both writes have completed, so it will always be (4, 3). But single-assignment rules out this program(although we could fake it by using a distinct IVar for each component of the pair, but with more sophisticated data structures it becomes difficult to fake a solution with IVars.)\par

@@

For LVars, the frame property says that independent effects commute with each other. Consider an expression $e$ that runs starting in \emph{store S} and steps to the expression $e{′}$, updating the \emph{store} to $S{′}$. Frame property guarantees that if $e$ starts from a larger $store S$ $\sqcup$ $R$, where $R$ is some other store \emph{“framing on”} to $S$, then $e$  will update the $store$ to $S′$ $\sqcup$ $R$, and $e$ will step to $e′$. The \emph{“frame”} \emph{store} $R$  is a \emph{store} resulting from some other independently-running computation. $S$ and $R$ (or $S′$ and $R$) need not be disjoint — we are combining them with $\sqcup$, not with $∗$. If they did have to be disjoint, it would mean that concurrent operations cannot touch the same parts of the store. For LVars, this kind of total disjointness is unnecessary, since updates commute! Instead, we can behave as though $S$ and $R$ are disjoint, even though they might not be. \\ \\
\textbf{\textit{Diamond Lemma:}} This does the heavy lifting of the determinism proof: it establishes the diamond property, which says that if a configuration steps to two different configurations, there exists a single third configuration to which those configurations both step. For supporting lemmas, its details and complete proofs please refer to\cite{lkuper}. Lemmas there also show how cases of deadlocks and livelocks are also handled.

@@

In this regard, Lvish\cite{lkuper2} extends the Lvars model with 2 additional capabilities. First, it adds event handlers, a mechanism for attaching a callback function to an LVar that runs, asynchronously, whenever events arrive (in the form of monotonic updates to the Lvar). Second, a primitive for freezing an LVar, which means: once an LVar is frozen, any further writes will throw an exception; on the other hand, it becomes possible to discover the exact value of the LVar, learning both positive and negative information about it, without blocking.\par
Freezing does not commute with writes. If a freeze is interleaved before such a write, the write will raise an exception; if it is interleaved afterwards, the program will proceed normally. Although it appears like determinism is lost, these programs satisfy the property of Quasi determinism that is they will always output the same result or an error.

@@

This registers a handler for $lv$ that executes the callback function $\lambda{x}.put\ lv\ x + 1$ for each odd number that $lv$ is at or above. Event sets are a mathematical modeling tool only. Event handlers in LVish invoke their callback for all events in their event set Q that have taken place even if those events occurred prior to the handler being registered. If a callback only executed for events that arrived after its handler was registered, or only for the largest event in its handler set that had occurred, then it could result in non-determinism. For example if puts are interleaved with handler registration as a part of a $par$ expression then a $get$ operation outside would block, or not, depending on how the handler registration was interleaved with the $puts$. By instead executing a handlers callback once for each and every element in its event set below or at the LVars value, quasi-determinism is guaranteed. \\ \\

@@

The freeze operation is a non-blocking read that lets us find out the exact contents of an LVar. After freezing, any further writes that would change LVar's state will raise an exception! A safe practice is to quiece before freezing For a program that performs freezes, Only two possible outcomes are guaranteed: the deterministic result of all the effects, or a write-after-freeze exception that can assist in debugging the synchronization bug. This property is called quasi-determinism, and the Lvish model is
quasi-deterministic.

Freezing introduces quasi-determinism to the programming model, because we might freeze before we have quiesced, or because our thread that's freezing and quiescing might be racing with some other thread to write into a shared LVar. But, if it is ensured that freezes only ever happen after all writes have completed, then the computations will be deterministic, because there wouldn't be any write-after-freeze races. This is enforced at the implementation level which takes care of quiescing and
freezing at the end of a computation. In the LVish library, this is provided by the runParThenFreeze function.\\
\cite{lkuper2}contains a quasi-deterministic, parallel, call-by-value $\lambdaLVish$-calculus extended with a store containing LVars. It extends the $\lambdaLVar$ calculus to support event handlers and freezing. Rather than modeling the full ensemble of event handlers, handler pools, quiescence, and freezing as separate primitives, it instead formalizes the \emph{“freeze-after”} pattern—which is a simpler primitive combining all others.

An LVar’s state is now a pair $(d,frz )$, where $d$ is an element of the application-specific set $D$ and $frz$(Short for freeze) is a “status bit” of either $true$ or $false$. We define an ordering $\leqp$ on LVar states $\state{d}{\status}$ in terms of the application-specific ordering $\userleq$ on elements of $D$. Every element of $D$ is ``freezable'' except $\top$. Informally: 
\begin{itemize}
\item Two unfrozen states are ordered according to the application-specific $\userleq$; that is, $(d,false) \leqp (d',false)$
exactly when $d \userleq d'$.
\item Two frozen states do not have an order, unless they are equal: $(d,true) \leqp (d',true)$ exactly when $d = d'$.
\item An unfrozen state $(d,false)$ is less than or equal to a frozen state $(d',true)$ exactly when $d \userleq d'$.
\item A frozen state is less than an unfrozen state if the unfrozen state is $\top$; that is, $(d,true) \leqp (d',false)$ exactly when $d' = \top$.
\end{itemize}
The addition of status bits to the application-specific lattice results in a new lattice $(D_p, \leqp, \botp, \topp)$. $\lubp{}{}$ for the lub operation that $\leqp$ induces. During the evaluation of LVish programs, a \emph{store} $S$ keeps track of the states of LVars and represents them by a binding from a location $l$, drawn from a set $loc$, to its state, which is some pair $(d, frz)$ from the set $D_p$. For detailed syntax and semantics of Lvish please refer to the paper\cite{lkuper2}.\\ \\
\textbf{\textit{Communication primitives:}} They are same as Lvar except that new will now create a binding for a new Lvar whose initial state is ($\bot$, false). $Put$ takes a pointer to the Lvar and a new lattice element which is a pair. $Get$ similarly takes a threshold set containing pairs. $freeze\ e_{lv}\ after\ e_{events}\ with\ e_{cb}$ has the following semantics: It attaches the callback $e_{cb}$ to the LVar $e_{lv}$. The expression $e_{events}$ must evaluate to a event set $Q$; the callback will be executed, once, for each lattice element in $Q$ that the LVar’s state reaches or surpasses. The callback takes a lattice element as its argument. Its return value is ignored, so it runs solely for effect. For instance, a callback might itself do a put to the LVar to which it is attached, triggering yet more callbacks. If the handler reaches a quiescent state, the LVar $e_{lv}$ is frozen, and its exact state is returned (rather than an under-approximation of the state, as with $get$).

@@

It adopts the basic approach of the par monad. The existing Par monad\cite{simon} in Haskell offers an API for parallel programming. It provides by default a work-stealing scheduler and supports forking tasks. The programmer specifies how information flows across computations but not the order in which they will be evaluated at runtime. Information flow is described using IVars. With
LVish we get a lattice-generic infrastructure: the Par monad itself, a thread scheduler, support for blocking and signaling threads, handler pools, and event handlers. This infrastructure does not guarantee quasi-determinism, only data structure
authors should import it who will implement a specific monotonic data structure using this library providing efficient parallel access, exporting a limited interface for its users ensuring quasi-determinism. LVish library provides its own custom scheduler and programs run only inside the par monad allowing only Lvish-sanctioned side effects to provide compile time guarantees of

@@

In order to specify the type of the par computation it provides a phantom type that indicates the determinism level.

@@

Par : : Determinism -> * -> *

@@

runPar : : Par Det a -> a
runParIO : : Par l v l a -> IO a
runParThenFreeze : : Deep frz a =>

>>>>

The purpose of a deterministic-by-construction programming model is to rule out programs like the following from being written, or force them to behave deterministically. \cite{lindsey}Written in a hypothetical parallel language. 
\begin{verbatim}
let _ = put lv 3 in  /write 3 in lv/
let par v = get lv  /read from lv/
        _ = put lv 4  /write 4 in lv/
	      in v   /final value is v/
\end{verbatim} 
The \textit{let par} expression which has \textit{fork-join semantics}, evaluates the two sub-expressions in parallel. The value of $v$ is the value of the entire program. Depending on the scheduler \textit{get} $lv$ or \textit{put} $lv\ 4$ can run first, $v$ might be either $3$ or $4$ during multiple runs giving rise to non determinism.\par
A popular approach for enforcing determinism is to require that variables can only be assigned to once, giving us a single-assignment language. Our program would be invalid because it tries to write to $lv$ twice. Single-assignment variables with blocking read semantics, often known as IVars, turn up in all kinds of deterministic parallel systems. But the single-assignment rule disallows a number of programs that we would like to permit such as writing the same value to an IVar twice. As another example, consider the problem of filling in various parts of a data structure in parallel. \cite{lindsey}Say that $lv$ contains a pair, and we are writing to $lv$ twice, with each write filling in one component of the pair. Please note that $x=\bot$ below.\par 
\begin{verbatim}
let par  _ = put lv (4,x)
         _ = put lv (x,3) /error/
         in get lv
\end{verbatim}  
Read from $lv$ is possible once both writes have completed, so it will always be (4, 3). But single-assignment rules out this program(although we could fake it by using a distinct IVar for each component of the pair, but with more sophisticated data structures it becomes difficult to fake a solution with IVars.)\par

@@

For LVars, the frame property says that independent effects commute with each other. Consider an expression $e$ that runs starting in \emph{store S} and steps to the expression $e^{'}$, updating the \emph{store} to $S^{'}$. Frame property guarantees that if $e$ starts from a larger $store\ S$ $\sqcup$ $R$, where $R$ is some other $store$ \emph{“framing on”} to $S$, then $e$  will update the $store$ to $S^{′}$ $\sqcup$ $R$, and $e$ will step to $e^{′}$. The \emph{“frame”} \emph{store} $R$  is a \emph{store} resulting from some other independently-running computation. $S$ and $R$ (or $S^{′}$ and $R$) need not be disjoint, we are combining them with $\sqcup$, not with $\ast$. If they did have to be disjoint, it would mean that concurrent operations cannot touch the same parts of the store. For LVars, this kind of total disjointness is unnecessary, since updates commute! Instead, we can behave as though $S$ and $R$ are disjoint, even though they might not be. \\ \\
\textbf{\textit{Diamond Lemma:}} This does the heavy lifting of the determinism proof: it establishes the diamond property, which says that if a configuration steps to two different configurations, there exists a single third configuration to which those configurations both step. For supporting lemmas, its details and complete proofs please refer to\cite{lkuper}. Lemmas there also show how cases of deadlocks and livelocks are also handled.

@@

Lvish\cite{lkuper2} extends the Lvars model with 2 additional capabilities. First, it adds event handlers, a mechanism for attaching a callback function to an LVar that runs, asynchronously, whenever events arrive (in the form of monotonic updates to the Lvar). Second, a primitive for freezing an LVar, which means: once an LVar is frozen, any further writes will throw an exception; on the other hand, it becomes possible to discover the exact value of the LVar, learning both positive and negative information about it, without blocking.\par
Freezing does not commute with writes. If a freeze is interleaved before such a write, the write will raise an exception; if it is interleaved afterwards, the program will proceed normally. Although it appears like determinism is lost, these programs still satisfy the property of Quasi determinism that is they will always output the same result or an error.

@@

This registers a handler for $lv$ that executes the callback function $\lambda{x}.put\ lv\ x + 1$ for each odd number that $lv$ is at or above. Event sets are a mathematical modeling tool only. Event handlers in LVish invoke their callback for all events in their event set Q that have taken place even if those events occurred prior to the handler being registered. If a callback only executed for events that arrived after its handler was registered, or only for the largest event in its handler set that had occurred, then it could result in non-determinism. For example if $puts$ are interleaved with handler registration as a part of a $par$ expression then a $get$ operation outside would block, or not, depending on how the handler registration was interleaved with the $puts$. By instead executing a handlers callback once for each and every element in its event set below or at the LVars value, quasi-determinism is guaranteed. \\ \\

@@

The freeze operation is a non-blocking read that lets us find out the exact contents of an LVar. After freezing, any further writes that would change LVar's state will raise an exception! A safe practice is to quiece before freezing For a program that performs freeze, Only two possible outcomes are guaranteed: the deterministic result of all the effects, or a write-after-freeze exception that can assist in debugging the synchronization bug. This property is called quasi-determinism, and the Lvish model is
quasi-deterministic.

Freezing introduces quasi-determinism to the programming model, because we might freeze before we have quiesced, or because our thread that's freezing and quiescing might be racing with some other thread to write into a shared LVar. But, if it is ensured that freezes only ever happen after all writes have completed, then the computations will be deterministic, because there wouldn't be any write-after-freeze races. This is enforced at the implementation level which takes care of quiescing and
freezing at the end of a computation. In the LVish library, this is provided by the runParThenFreeze function.\\
\cite{lkuper2}contains a quasi-deterministic, parallel, call-by-value $\lambdaLVish$-calculus extended with a store containing LVars. It extends the $\lambdaLVar$ calculus to support event handlers and freezing. Rather than modeling the full ensemble of event handlers, handler pools, quiescence, and freezing as separate primitives, it instead formalizes the \emph{“freeze-after”} pattern which is a simpler primitive combining all others.

An LVar’s state is now a pair $(d,frz )$, where $d$ is an element of the application-specific set $D$ and $frz$(Short for freeze) is a “status bit” of either $true$ or $false$. We define an ordering $\leqp$ on LVar states $\state{d}{\status}$ in terms of the application-specific ordering $\userleq$ on elements of $D$. Every element of $D$ is ``freezable'' except $\top$. Informally: 
\begin{itemize}
\item Two unfrozen states are ordered according to the application-specific $\userleq$; that is, $(d,false) \leqp (d',false)$
exactly when $d \userleq d'$.
\item Two frozen states do not have an order, unless they are equal: $(d,true) \leqp (d',true)$ exactly when $d = d'$.
\item An unfrozen state $(d,false)$ is less than or equal to a frozen state $(d',true)$ exactly when $d \userleq d'$.
\item A frozen state is less than an unfrozen state if the unfrozen state is $\top$; that is, $(d,true) \leqp (d',false)$ exactly when $d' = \top$.
\end{itemize}
The addition of status bits to the application-specific lattice results in a new lattice $(D_p, \leqp, \botp, \topp)$. $\lubp{}{}$ for the lub operation that $\leqp$ induces. During the evaluation of LVish programs, a \emph{store} $S$ keeps track of the states of LVars and represents them by a binding from a location $l$, drawn from a set $loc$, to its state, which is some pair $(d, frz)$ from the set $D_p$. For detailed syntax and semantics of Lvish please refer to the paper\cite{lkuper2}.\\ \\
\textbf{\textit{Communication primitives:}} They are same as Lvar except that $new$ will now create a binding for a new Lvar whose initial state is ($\bot$, false). $put$ takes a pointer to the Lvar and a new lattice element which is a pair. $get$ similarly takes a threshold set containing pairs. $freeze\ e_{lv}\ after\ e_{events}\ with\ e_{cb}$ has the following semantics: It attaches the callback $e_{cb}$ to the LVar $e_{lv}$. The expression $e_{events}$ must evaluate to a event set $Q$; the callback will be executed, once, for each lattice element in $Q$ that the LVar’s state reaches or surpasses. The callback takes a lattice element as its argument. Its return value is ignored, so it runs solely for effect. For instance, a callback might itself do a $put$ to the LVar to which it is attached, triggering yet more callbacks. If the handler reaches a quiescent state, the LVar $e_{lv}$ is frozen, and its exact state is returned (rather than an under-approximation of the state, as with $get$).

@@

It adopts the basic approach of the par monad. The existing $par$ monad\cite{simon} in Haskell offers an API for parallel programming. It provides by default a work-stealing scheduler and supports forking tasks. The programmer specifies how information flows across computations but not the order in which they will be evaluated at runtime. Information flow is described using IVars. With
LVish we get a lattice-generic infrastructure: the $par$ monad itself, a thread scheduler, support for blocking and signaling threads, handler pools, and event handlers. This infrastructure does not guarantee quasi-determinism, only data structure
authors should import it who will implement a specific monotonic data structure using this library providing efficient parallel access, exporting a limited interface for its users ensuring quasi-determinism. LVish library provides its own custom scheduler and programs run only inside the $par$ monad allowing only Lvish-sanctioned side effects to provide compile time guarantees of

@@

In order to specify the type of the $par$ computation it provides a phantom type that indicates the determinism level.

@@

Par :: Determinism -> * -> *

@@

runPar :: Par Det a -> a
runParIO :: Par l v l a -> IO a
runParThenFreeze :: Deep frz a =>
